Supporting Dynamic Data Partitioning in Parallel Programs

### 2012-11-03 doc/types-pldi2013/example.tex
edits to intro, saving space

% This is a description of how the listings should be formatted.
% It can go anywhere before the listings.
\lstset{
  captionpos=b,
  language=Haskell,
  basicstyle=\scriptsize,
  numbers=left,
  numberstyle=\tiny,
  columns=fullflexible,
  stepnumber=1,
  escapechar=\#,
  keepspaces=true,
  literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
  morekeywords={function,rr,int,float,bool,isnull,partition,as,downregion,upregion,reads,writes,rdwrs,reduces,read,write,reduce,using,unpack,pack,coloring,multicoloring,color,newcolor,atomic,simultaneous},
  deletekeywords={float,head,min,max}
}

>>>>

% This is a description of how the listings should be formatted.
% It can go anywhere before the listings.
\lstset{
  captionpos=b,
  language=Haskell,
  basicstyle=\scriptsize,
  numbers=left,
  numberstyle=\tiny,
  columns=fullflexible,
  stepnumber=1,
  escapechar=\#,
  keepspaces=true,
  belowskip=-10pt,
  literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
  morekeywords={function,rr,int,float,bool,isnull,partition,as,downregion,upregion,reads,writes,rdwrs,reduces,read,write,reduce,using,unpack,pack,coloring,multicoloring,color,newcolor,atomic,simultaneous},
  deletekeywords={float,head,min,max}
}

### 2012-11-03 doc/types-pldi2013/intro.tex
edits to intro, saving space

Previous work has explored two approaches for describing the organization of data
in distributed memory parallel programs.

One class of work uses fully static analyses with no runtime overhead, but must
disallow potential data aliasing to make the analyis tractable, limiting expressivity.
Two recent examples, Sequoia \cite{Fatahalian06} and 
Deterministic Parallel Java (DPJ) \cite{Bocchino09}, each provide a 
mechanism to statically partition the heap.
% into a tree of collections of data.
The two designs are different in many aspects, but agree that there is
a single tree-shaped partitioning of data that must be checked statically 
(see Section~\ref{sec:related}).  The second approach permits potential data
aliasing, allowing greater expressivity, but must fully dynamic analyses of
program execution to detect aliasing situations and guarantee correct results.
However, this approach incurs significant runtime overhead and requires 
centralized control logic that is difficult to scale to distributed 
memory machines; examples include transactional memory \cite{Harris05} 
and thread-level speculation \cite{Steffan00}.

@@

\item  a logical region may be dynamically partitioned in multiple different ways.
Subregions from multiple partitions may include the same data, introducing aliasing.

@@

To efficiently guarantee the safety and
correctness of programs using first-class regions,
privilege and coherence properties in the presence of arbitrary region
aliasing, Legion relies on a combination of static and
dynamic checks.  A key theme that runs through the design is that 
alias analysis is a very hard static analysis problem, but is both easy and
inexpensive if they done dynamically at the granularity of regions instead of individual
heap locations, while checks that must be done on individual values are only
cheap if done statically.  Explicating and evaluating the design of Legion's
static and dynamic semantics and its implications are the topics of this paper,
which we present as a series of contributions:
\begin{itemize}
\item We present a type system for {\em Core Legion} programs that statically
  verifies the safety of individual pointers and region
  privileges at task call boundaries (Section~\ref{subsec:coretypes}).
%  A subtle issue arises in checking privileges when regions are
%  stored in and later retrieved from heap data structures.  The type
%  system provides enough static information that the Legion runtime
%  can leverage its dynamic information about region aliasing to
%  perform dynamic privilege checks guaranteeing safety.

@@

% Putting this here so that we can get the code in a good place
\lstset{
  captionpos=b,
  language=Haskell,
  basicstyle=\scriptsize,
  numbers=left,
  numberstyle=\tiny,
  columns=fullflexible,
  stepnumber=1,
  escapechar=\#,
  keepspaces=true,
  literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
  morekeywords={function,rr,int,float,bool,isnull,partition,as,downregion,upregion,reads,writes,rdwrs,reduces,read,write,reduce,using,unpack,pack,coloring,multicoloring,color,newcolor,atomic,simultaneous},
  deletekeywords={float,head,min,max}
}
\begin{lstlisting}[float={t},label={lst:circuit_ex},caption={Circuit Simulation}]
--                        <voltage,current,charge,capacitance>
type CircuitNode        = <float,float,float,float>
--                     < owned node, owned or ghost node, resistance, current>
type CircuitWire<rn,rg>  = <CircuitNode@rn, CircuitNode@(rn,rg),float,float>

>>>>

Previous work has explored two approaches for describing the organization of data
in distributed memory parallel programs.
One class of work uses fully static analyses with no runtime overhead, but 
disallows potential data aliasing to make the analyis tractable, limiting expressivity.
Two recent examples, Sequoia \cite{Fatahalian06} and 
Deterministic Parallel Java (DPJ) \cite{Bocchino09}, each provide a 
mechanism to statically partition the heap.
% into a tree of collections of data.
The two designs are different in many aspects, but agree that there is
a single tree-shaped partitioning of data that must be checked statically 
(see Section~\ref{sec:related}).  The second approach permits potential data
aliasing, allowing greater expressivity, but uses fully dynamic analyses of
program execution to detect aliasing situations and guarantee correct results.
This approach incurs significant runtime overhead and requires 
centralized control logic that is difficult to scale to distributed 
memory machines; examples include transactional memory \cite{Harris05} 
and thread-level speculation \cite{Steffan00}.

@@

\item  a logical region may be dynamically partitioned in multiple different ways;
subregions from multiple partitions may include the same data, introducing aliasing.

@@

To guarantee the safety and correctness of programs, Legion relies on
a combination of static and dynamic analysis based on the privilege
and coherence systems.  
%To efficiently guarantee the safety and
%correctness of programs using first-class regions,
%privilege and coherence properties in the presence of arbitrary region
%aliasing, Legion relies on a combination of static and
%dynamic checks.  
The critical feature in Legion that makes static analysis tractable
and dynamic analysis efficient in the presence of potential aliasing is logical regions.
Alias analysis is a very hard static analysis problem, but is both easy and
inexpensive if done dynamically at the granulatiry of logical regions instead
of individual memory locations; similarly
checks on individual memory accesses are expensive if done dynamically, but
cheap if done statically within the context of local logical region privileges.
%A key theme that runs through the design is that 
%alias analysis is a very hard static analysis problem, but is both easy and
%inexpensive if done dynamically at the granularity of regions instead of individual
%heap locations, while checks that must be done on individual values are only
%cheap if done statically.  
Explicating and evaluating the design of Legion's
static and dynamic semantics and its implications are the topics of this paper,
which we present as a series of contributions:
\begin{itemize}
\item We present a type system for {\em Core Legion} programs that statically
  verifies the safety of individual pointers and region
  privileges at task call boundaries (Section~\ref{subsec:coretypes}).
%  A subtle issue arises in checking privileges when regions are
%  stored in and later retrieved from heap data structures.  The type
%  system provides enough static information that the Legion runtime
%  can leverage its dynamic information about region aliasing to
%  perform dynamic privilege checks guaranteeing safety.

@@

% Putting this here so that we can get the code in a good place
\lstset{
  captionpos=b,
  language=Haskell,
  basicstyle=\scriptsize,
  numbers=left,
  numberstyle=\tiny,
  columns=fullflexible,
  stepnumber=1,
  escapechar=\#,
  keepspaces=true,
  belowskip=-10pt,
  literate={<}{{$\langle$}}1 {>}{{$\rangle$}}1,
  morekeywords={function,rr,int,float,bool,isnull,partition,as,downregion,upregion,reads,writes,rdwrs,reduces,read,write,reduce,using,unpack,pack,coloring,multicoloring,color,newcolor,atomic,simultaneous},
  deletekeywords={float,head,min,max}
}
\begin{lstlisting}[float={t},label={lst:circuit_ex},caption={Circuit Simulation}]
--                        <voltage,current,charge,capacitance>
type CircuitNode        = <float,float,float,float>
--                     < owned node, owned or ghost node, resistance, current>
type CircuitWire<rn,rg>  = <CircuitNode@rn, CircuitNode@(rn,rg),float,float>
